---
title: "Activity: Bird Species"
author: "Your Names Here"
date: "DUE DATE "
output: html_notebook
---


# Set Up

```{r}
# Load `tidyverse` and `dcData` packages

library(tidyverse)
library(dcData)


# Load the `OrdwayBirds` data set into your RStudio environment from the `dcData` package

data(OrdwayBirds)



# Show a few commands used to inspect the data set a few different ways

str(OrdwayBirds)
summary(OrdwayBirds)

```

# Activity

When you're finished with this activity, you will have a graph that shows what time of year various species appear at the Katherine Ordway Natural History Study Area in Inver Grove Heights, MN.

### Step 0

Before we begin, the book gives us some instructions to select a few key variables and clean up the date formatting.  Use the commands given in the book, but be sure to study the functions used so you know what is happening.  In this case, we select `SpeciesName`, `Month`, and `Day`. Then a `mutate()` command converts each variable to character, and then converts the resulting "character string" to a number.

```{r}
# Get the data table & clean up dates (see Data Computing eBook)

OrdwayBirds <-
  OrdwayBirds %>%
  select(SpeciesName,Month,Day) %>%
  mutate(Month= as.numeric(as.character(Month)),Day=as.numeric(as.character(Day)))
  

# Inspect resulting data


```



### Step 1




```{r}
# unique species in original data (including mis-spellings)

distinct_species <- n_distinct(OrdwayBirds$SpeciesName)


# unique species in the clean list

distinct_species_cleaned <- n_distinct(OrdwaySpeciesNames$SpeciesNameCleaned)


```


### Step 2
  



```{r}
# inspect OrdwaySpeciesNames
Corrected <- OrdwayBirds %>%
 
  

# inner join

inner_join(OrdwaySpeciesNames) %>%
  select(Species= SpeciesNameCleaned, Month, Day) %>% na.omit()
head(Corrected)

# inspect the data after the join


```


### Step 3
We want to identify the top 5-6 bird species sighted at the Katherine Ordway Natural History Study Area.

```{r}
# count sightings by species in descending order

sightings <- Corrected %>%
  count(Species, sort = TRUE)

top_species <- head(sightings, n = 6)



# a hint in the text book recommends displaying the top 10 species to choose our threshold

top_species <- Corrected %>%
  count(Species, sort = TRUE) %>%
  head(10)



# define a sightings threshold; produce a table with only birds belonging to major species

threshold <- 50  

major_species <- Corrected %>%
  count(Species, sort = TRUE) %>%
  filter(n > threshold)

MajorBirds <- Corrected %>%
  filter(Species %in% major_species$Species)



```


### Step 4

We now want to produce a month-by-month count for each of the major species, and then construct a chart to tell the story of what time of year the various species appear.

```{r}
# prepare glyph-ready data table called `ByMonth`

ByMonth <- MajorBirds %>%
  group_by(Month, Species) %>%
  summarise(Count = n(), .groups = "drop")


```

Now use the ggplot template below to generate the desired graph. (Remark you can also use esquisser or mplot to create a line chart instead of the barchart)



```{r}
 ByMonth %>%
  ggplot(aes(x = Month, y = Count, color = Species)) +
  geom_line()


```

According to the graph, we can answer the questions posed:

##### 1. Which species are present year-round?  
"American Goldfinch"      "Black-capped Chickadee"  "White-breasted Nuthatch"


##### 2. Which species are migratory, that is, primarily present in one or two seasons?

"Black-throat Sparrow"


##### 3. What is the peak month for each major species? (bullet list)

[1] "10" "5"  "11" "1"  "3"  "5"  "8"  "9"  "5"  "5"  "7"  "5"  "10" "10" "9"  "10" "9"  "10"
[19] "5"  "5"  "5"  "10" "10" "4"  "5"  "5"  "9"  "3"  "10" "10" "10" "5" 


##### 4. Which major species are seen in good numbers for at least 6 months of the year?

[1] "American Goldfinch"      "Black-capped Chickadee"  "Cardinal"               
 [4] "Catbird"                 "Cedar Waxwing"           "Chipping Sparrow"       
 [7] "Field Sparrow"           "House Wren"              "Least Flycatcher"       
[10] "Lincoln's Sparrow"       "Myrtle Warbler"          "Purple Finch"           
[13] "Red-Winged Blackbird"    "Robin"                   "Rose Breasted Grosbeak" 
[16] "Slate-colored Junco"     "Song Sparrow"            "Tree Swallow"           
[19] "White-breasted Nuthatch"

```{r}
year_round_species <- ByMonth %>%
  group_by(Species) %>%
  summarise(Num_Months = n_distinct(Month))

year_round_species <- year_round_species %>%
  filter(Num_Months == 12)

year_round_species$Species

migratory_species <- ByMonth %>%
  group_by(Species) %>%
  summarise(Num_Months = n_distinct(Month))

migratory_species <- migratory_species %>%
  filter(Num_Months <= 2)

migratory_species$Species

peak_months <- ByMonth %>%
  group_by(Species) %>%
  filter(Count == max(Count)) %>%
  summarise(Peak_Months = paste(Month, collapse = ", "))

peak_months$Species
peak_months$Peak_Months

species_six_months <- ByMonth %>%
  group_by(Species) %>%
  summarise(Num_Months = n_distinct(Month)) %>%
  filter(Num_Months >= 6)

species_six_months$Species




```

